Statically Typed String Sanitation Inside a Python

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex
Added check rules; lam_rs and lam_p are now full systems.
still need to proof read

@@

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x {:} \sigma . e$ \alt
      $e~(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\coerce{r}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
%\inferrule[S-T-Abs]
%{\sctx, x : T_1 \vdash e : T_2}
%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[S-T-App]
%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
%{\sctx \vdash e_1 e_2 : T}
%  
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

\inferrule[S-T-coerce]
{\sctx \vdash e : \stringin{r'}}
{\sctx \vdash \coerce{r}{e} : \stringin{r}}

@@

\caption{Typing rules for our fragment of $\lambdas$. 
The typing context $\sctx$ is standard.}
\label{fig:etyping}
\end{figure}

\begin{figure}[t]
\small
$\fbox{\inferrule{}{S \sreduces S}}$
$\fbox{\inferrule{}{S \err}}$
\begin{mathpar}
%
%\inferrule[S-E-Abs]                                                             
%{ \ }                                                                           
%{\lambda x.e \sreduces \lambda x.e}                                             
%
%\inferrule[S-E-App]
%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
%{ e_1 e_2 \sreduces e_1' e_2'}
%
\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

@@

\inferrule[S-E-Coerce-OK]
{S \sreduces \strin{s} \\ s \in \lang{r}}
{\coerce{r}{S} \sreduces \strin{s}}

\end{mathpar}
\caption{Big step semantics for our fragment of $\lambdas$.
Error propagation rules are omitted.}
\label{fig:eeval}
\end{figure}

@@

\begin{mathpar}
%\inferrule[P-T-Abs]
%{\tctx, x : T_1 \vdash e : T_2}
%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[P-T-App]
%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
%{\tctx \vdash e_1 e_2 : T}
%
\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

@@

\inferrule[P-T-Check]
{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ }
{\tctx \vdash \tcheck{\iota_1}{\iota_2} : \str}
\end{mathpar}
\caption{Typing rules for our fragment of $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:ityping}
\end{figure}

@@

\begin{mathpar}

%\inferrule[P-E-Abs]
%{ \ }
%{\lambda x.e \sreduces \lambda x.e}
%
%\inferrule[P-E-App]
%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
%{ e_1 e_2 \sreduces e_1' e_2'}
%
  \inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

@@

\inferrule[P-E-Check-Err]
{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
{\tcheck{\iota_1}{\iota_2} \err}
\end{mathpar}
\caption{Big step semantics for our fragment of $\lambdap$.
Error propagation rules are omitted.}
\label{fig:ieval}
\end{figure}

@@

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types; an explanation of significant typing rule follows:
\begin{itemize}
  \item Rule S-T-Concat states that the type of concatenated strings is 
    obtained by concatenating the regular expressions for each string.
  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item Rule S-T-Coerce allows coercion; however, note that this coercion cannot
    invalidate our safety property because we ensure appropraite checks are always
    inserted wherever coercions are used.
\end{itemize}

@@

Safety for the string fragment is more involved, because it involves validating that
the type system's definition is justified by our theorems about regular languages.
Note that type safety does \emph{not} guarantee that strings of a type are in the
correct language. We isolate this property so that we may reason about its preservation
under translation to $\lambdap$, which does not have regular expression types.

@@

\begin{thm}{Translation Correctness.} \label{thm:tcorrect}
If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$, or
\end{thm}

>>>>

\newcommand{\rcheck}[4]{ {\sf rcheck}[#1](#2;#3;#4) }

@@

<$e$> ::= 
      $x$ \hfill source terms \alt 
      $\lambda x {:} \sigma . e$ \alt
      $e~(e)$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{e}{e}$ \alt
      $\sisubst{r}{e}{e}$ \alt
      $\coerce{r}{e}{x.e}{e}$
\caption{Syntax of $\lambda_{RS}$.}\label{fig:lamrs}
\end{grammar}
\end{figure}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
\inferrule[S-T-Abs]
{\sctx, x : \sigma_1 \vdash e : \sigma_2}
{\sctx \vdash \lambda x.e : \sigma_1 \rightarrow \sigma_2}

\inferrule[S-T-App]
{\sctx \vdash e_1 : \sigma_1 \\ \sctx \vdash e_2 : \sigma_1 \rightarrow \sigma}
{\sctx \vdash e_1 e_2 : \sigma}

\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

\inferrule[S-T-SafeCoerce]
{\sctx \vdash e : \stringin{r'} \\ \lang{r'} \subseteq \lang{r}}
{\sctx \vdash \coerce{r}{e} : \stringin{r}}

\inferrule[S-T-Check]
{\sctx \vdash e_0 : \stringin{r_0} \\ \sctx, x:\stringin{r} \vdash e_1 : \sigma \\ \sctx \vdash e_2 : \sigma}
{\sctx \vdash \rcheck{r}{e_0}{x.e_1}{e_2} : \sigma}

@@

\caption{Typing rules for $\lambdas$. 
The typing context $\sctx$ is standard.}
\label{fig:etyping}
\end{figure}

\begin{figure}[t]
\small
$\fbox{\inferrule{}{S \sreduces S}}$
$\fbox{\inferrule{}{S \err}}$
\begin{mathpar}
\inferrule[S-E-Abs]                                                             
{ \ }                                                                           
{\lambda x.e \sreduces \lambda x.e}

\inferrule[S-E-App]
{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
{ e_1 e_2 \sreduces e_1' e_2'}

\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

@@

\inferrule[S-E-SafeCoerce]
{e \sreduces \strin{s}}
{\coerce{r}{e} \sreduces \strin{s}}

\inferrule[S-E-Check-Ok]
{e \sreduces \strin{s} \\ s \in \lang{r}}
{
  \rcheck{r}{e}{x.e_1}{e_2} \sreduces [\strin{s} / x]e_1
}

\inferrule[S-E-Check-NotOk]
{
  e \sreduces \strin{s} \\ s \not \in \lang{r}
}
{
  \rcheck{r}{e}{x.e_1}{e_2} \sreduces e_2
}

\end{mathpar}
\caption{Big step semantics for $\lambdas$.
Error propagation rules are omitted.}
\label{fig:eeval}
\end{figure}

@@

\begin{mathpar}
\inferrule[P-T-Abs]
{\tctx, x : \iota_1 \vdash e : \iota_2}
{\tctx \vdash \lambda x.e : \iota_1 \rightarrow \iota_2}

\inferrule[P-T-App]
{\tctx \vdash e_1 : \iota_1 \\ \tctx \vdash e_2 : \iota_1 \rightarrow \iota}
{\tctx \vdash e_1 e_2 : \iota}

\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

@@

\inferrule[P-T-Check]
{\tctx \vdash \iota_1 : \regex \\ \tctx \vdash \iota_2 : \str \\ }
{\tctx \vdash \tcheck{\iota_1}{\iota_2} : \str}
\end{mathpar}
\caption{Typing rules for $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:ityping}
\end{figure}

@@

\begin{mathpar}
\inferrule[P-E-Abs]
{ \ }
{\lambda x.e \sreduces \lambda x.e}

\inferrule[P-E-App]
{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
{ e_1 e_2 \sreduces e_1' e_2'}

  \inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

@@

\inferrule[P-E-Check-Err]
{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \tstr{s} \\ s \not \in \lang{r}}
{\tcheck{\iota_1}{\iota_2} \err}
\end{mathpar}
\caption{Big step semantics for of $\lambdap$.
Error propagation rules are omitted.}
\label{fig:ieval}
\end{figure}

@@

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types; an explanation of significant typing rule follows:
\begin{itemize}
  \item Rule S-T-Concat states that the type of concatenated strings is 
    obtained by concatenating the regular expressions for each string.
  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item Rule S-T-SafeCoerce allows only safe coercions between string types by
    expoiting the decidability of language inclusion.
\end{itemize}

@@

Safety for the string fragment of $\lambdas$ is more involved because it involves validating that
the type system's definition is justified by our theorems about regular languages.
We state the theorem for the string fragment only to demontrate this point;
full type safety for $\lambdas$ follows easily from safety for the string fagment.
Note that type safety does \emph{not} guarantee that strings of a type are in the
correct language. We isolate this property so that we may reason about its preservation
under translation to $\lambdap$, which does not have regular expression types.

@@

\begin{thm}{Translation Correctness.} \label{thm:tcorrect}
If $\sctx \vdash e : \stringin{r}$ then there exists a $\iota$ such that $\trden{e} = \iota$, 
$\iota \treduces \tstr{s}$, and $e \sreduces \strin{s}$.
\end{thm}
